---
title: "Time Series Lab 2"
author: "Montsie Guerrero"
date: "10/27/2019"
output: html_document
---

   
```{r}
library(plm)
library(plyr)
library(stargazer)
pan=read.csv("panel-for-R.csv")
```
## 1. (a) Run an OLS regression, including at least one independent variable and a time variable (as dummies).  Explain how you think your independent variable relates to your dependent variable.  Interpret your results.  Did you find what you expected to find? 

Source of Data: [GSS Panel Data](https://gss.norc.org/)
I wanted to look at the effect on how a respondent's view on how important a child should learn obedience may be affected by his/her view on spanking a child as a form of discipline. The dependent variable I investigated is **obey** which asks the question: Which would you pick as the most important for a child to learn to prepare him or her for life? To obey. This was rated on a scale of 1-5 (and recoded): Most Important (5), Second Important (4), Third Important (3), Fourth Important (2), and Least Important (1). The independent variable is **spanking**, recoded as a dummy variable, and asks the question: Do you strongly agree, agree, disagree or strongly disagree that it is sometimes necessary to discipline a child with a good, hard, spanking?. Agree (1) and Disagree (0). I hypothesize that those who believe strongly that spanking can discipline children would have a positive correlation with the view that obedience is important for children to learn. This is because historically, spanking has been used as positive punishment by parents who want to teach their children to behave appropriately. Obedience is a virtue that requires following certain norms, manners, and forms of authority and is brought about through discipline.

The data is collected in 3 panel waves: 2006, 2008, and 2010. This data is unbalanced because there are missing data.

```{r}
vars = c("spanking", "obey", "idnum", "panelwave", "childs")
lab = pan[,vars]

lab$spank1 = 5-lab$spanking
lab$spank = ifelse(lab$spank1>2, 1,0)
lab$obeychild = 6-lab$obey

```

*OLS Regression*
```{r}
ols.spank <- plm(obeychild ~ spank + as.factor(panelwave), # model formula
               index = c("idnum", "panelwave"), # id & time variables
               model = "pooling", 
               data = lab)

summary(ols.spank)
```

For a 1-unit increase in the view that spanking is a good disciplinary tactic, there is a 0.75 point increase in the view that children should learn obedience. The p-value of the coefficient on spank is highly statistically significant at the 0.001 level. We reject the null hypothesis. People in 2008, relative to those in 2006, decreased their view that children should learn obedience by -0.101 points, net of their view on spanking. The p-value is not statistically significant (>0.05) so we fail to reject the null hypothesis. People in 2010, relative to those in 2006, decreased their view that children should learn obedience by -0.074 points, net of their view on spanking. The p-value is not statistically significant (>0.05) so we fail to reject the null hypothesis. The adjusted R-squared is 0.06, which means the model explains 6% of the variation in the data.

This suggests that as the years go by, views on spanking may not play a role in the view on obedience based on this model. However, this may not be good enough because maybe those who already believe spanking is good for children held the same opinion on the importance of obedience regardless. Also, this model looks at people from 2006, 2008, and 2010 as totally different groups of people rather than the same person over the years. We may need to look into other kinds of models to see if there is a better fit to explain the data.



## (b) Then run a fixed effect model version of that OLS model.  Interpret your results.  Did you find what you expected to find?  Why?  Why not? 
```{r}
fe.spank <- plm(obeychild ~ spank + as.factor(panelwave), # model formula
             index = c("idnum", "panelwave"), # id & time variables
             model = "within", 
             data = lab)
summary(fe.spank)

```

For a one unit change in the respondent's view on spanking, there is a 0.12 point positive change in their view that children should be taught obedience, on average, net of any particular person, across the first 3 waves of the panel. Since fixed effects control for the same person across the 3 panel waves, it is assumed that the individual-specific effects are correlated with the variable obey. It seems that there is still a positive correlation between the view on spanking and the view on obedience, while controlling for fixed effects. Also unlike the OLS model, the coefficients for the second and third wave of panel data show a positive correlation. However, the value of the coefficient (although positive) decreases in 2008 and 2010 (from 0.116 to 0.076 then 0.075). So, this may suggest that while there is still a positive correlation with the views on spanking and obedience, the point change difference is smaller, while controlling for fixed effects. Perhaps respondents' view on spanking factors less on the view on obedience as time goes by.

The p-value of the F-statistic however is 0.52, greater than the 0.05 significance level. It is not statistically significant so we fail to reject the hypothesis that the observed and unobserved fixed effects are equal to 0.

## (c) Then include an additional predictor in your fixed effects model that you think might account for the initial relationship you found between your X and your Y.  What effect does that new independent variable have in your new regression?

I added the variable **childs** which asks respondents how many children they have ever had from 0-8 or more. I hypothesize that adding this variable would add a beter explanation to the fixed effect model. Respondents who may have more children may strongly believe that children should be obedient, based on their own experiences with having children.

```{r}
fe2.spank <- plm(obeychild ~ spank + childs + as.factor(panelwave), # model formula
             index = c("idnum", "panelwave"), # id & time variables
             model = "within", 
             data = lab)
summary(fe2.spank)
```

For a one unit change in the respondent's view on spanking, there is a 0.15 point positive change in their view that children should be taught obedience, on average, net of any particular person, across the first 3 waves of the panel and holding other variables constant. For a one unit change in the number of children a respondent has, there is a -0.09 point negative change in their view that children should be taught obedience, on average, net of any particular person, across the first 3 waves of the panel and holding all other variables constant. The coefficient of 2008 is 0.09 which shows that the view on obedience is higher than that of the base of 2006. This is also the same for the coefficient of 2010 which is 0.08.

The addition of the explanatory variable **childs** is surprising. There is a negative correlation between **childs** and **obey** while the correlation between spank and obey remains positive in the model and is actually a higher value (0.15 vs. 0.12 previously). The addition of **childs** suggests that perhaps parents who have more children put less importance on teaching children to be obedient.

The p-value of the F-statistic however is 0.36, greater than the 0.05 significance level. It is not statistically significant so we fail to reject the hypothesis that the observed and unobserved fixed effects are equal to 0.


## (d) Then run a random effects model equivalent to your fixed effects model in step (b).  Interpret the results.
```{r}
re.spank <- plm(obeychild ~ spank + as.factor(panelwave), # model formula
             index = c("idnum", "panelwave"), # id & time variables
             model = "random", 
             data = lab)
summary(re.spank)

```

For a one unit increase in the view on spanking, there is a 0.47 point increase in the view that children should learn obedience, net of time, adjusting for the fact that the same person is repeatedly answering the same survey. The coefficient on spank is almost 4 times greater than the one from the fixed effects model (0.48 vs. 0.12). This coefficient is also statistically significant (<0.001). We reject the null hypothesis that the coefficient is statistically different from 0. The adjusted R-squared is also 0.049 which explains the individual effects. It is interesting to see that in 2008, the coefficient drops to -0.0012 points lower than the base of 2006. And in 2010, the coefficient is 0.014, a positive point increase in the view that children should be taught obedience compared to the base of 2006. However, these values are not statistically significant at the 0.05 level.

The p-value of the F-statistic is statistically significant (<0.05). We reject the null hypothesis that all random effects are 0.

## (e) Run a Hausman test to compare your fixed effects and your random effects models.  What do you conclude? 
```{r, warning=FALSE}
phtest(fe.spank, re.spank) 

stargazer(ols.spank, fe.spank, re.spank,
          title="Regression Results", 
          align=TRUE, 
          dep.var.labels=c("Obey"), 
          covariate.labels=c("Spank","2008", "2010"),  
          no.space=TRUE, 
          column.labels=c("Pooled", "Fixed Effects", "Random Effects"), 
          dep.var.caption="", 
          model.numbers=FALSE,
          type = "text", omit = "Constant")

```

The Hausman test is used to choose whether it is better to use the random effects or the fixed effects model. Based on the results, the p-value is below the 0.05 level. This means that it is statistically significant. We can reject the null hypothesis that the coefficients for random and fixed effects are the same. Based on the Hausman test, it is the fixed effects model that should be used. The fixed effects model eliminates omitted variable bias because it measures changes within groups across time by including dummy variables for the missing characteristics. There may be other unobserved variables that can affect the view on the importance of teaching children obedience that need to be controlled thus the fixed effects model is best.
